|

### 2023-01-27 paper/paper.Rmd
grammarly changes

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the contribution of a given variable to a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the center of a slice, for example, to explore how structure changes in local subspaces. The Mathematica package as well as example notebooks are provided, which contain functions enabling the user to experiment with these new manual controls, with one specifically for exploring regions and boundaries produced by classification models. The advantage of Mathematica is its linear algebra capabilities, and interactive cursor location controls. Some limited implementation has also been made available in the R package \tt{tourr}. 
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}\doublespacing
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, \textcolor{blue}{methods which work for arbitrary dimensions have more general utility}. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional  data with low-dimensional projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. (Various videos of tours, e.g. @dataviewer, @gt-pp-video, on the @ASA22 illustrate this concept.) Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video [@PRIM9-video] illustrating the capabilities is available through \textcolor{blue}{the} video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

<!-- Manual tours -->
One such direction of work develops the ideas from PRIM-9, to provide manual control of a tour. @cook_manual_1997 describe controls for 1D (or 2D) projections, respectively in a 2D (or 3D) manipulation space, allowing the user to select any variable axis, and rotate it into, or out of, or around the projection through horizontal, vertical, oblique, radial or angular changes in value. @spyrison_spinifex_2020 refined this algorithm and implemented it to generate radial tour animation sequences.

<!-- Purpose -->
Manual controls are especially useful for assessing sensitivity of structure to particular elements of the projection. There are many places where it is useful. In exploratory data analysis, where one sees clusters in a projection, one may ask whether some variables can be removed from the projection without affecting the clustering. For interpreting models, one can reduce or increase a variable's contribution to examine the variable importance. Having the user interact with a projection is extremely valuable for understanding high-dimensional data. However, these algorithms have two problems: (1) the pre-processing of creating a manipulation space overly complicates the algorithm, (2) extending to higher dimensional control is difficult.

Another potentially useful manual control, is to allow the user to choose the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the position of the center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center. It would also allow the user to interactively decide on the thickness of the slice.

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods is illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a Mathematica [@Mathematica] package that is used for the application, and describes the interactive environment that would be desirable within R [@rref] as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and three Mathematica notebooks that can be used to reproduce the applications.

@@

The methods for updating component values in @cook_manual_1997 (and utilized in @spyrison_spinifex_2020) are prescribed primarily for a 2D projection, to take advantage of (then) newly developed 3D trackball controls made available for computer gaming. The first step was to construct a 3D manipulation space from a 2D projection. In this space, the coefficient of the controlled variable ranges between -1 and 1. Movements of a cursor are recorded and converted into changes in the values of $V_m$ thus changing the displayed 2D projection. The algorithm also provided constraints to horizontal, vertical, radial or angular motions only. The construction of the manipulation space overly complicates the manual controls, especially when considering possible techniques that will apply to arbitrary $d$.

<!--
\begin{figure*}[ht]
\centerline{\includegraphics[width=0.8\textwidth]{figures/manip_space.pdf}}
\caption{Original contruction of the manual tour designed for 2D projections and created a 3D space from which to utilise track ball controls to change it's contribution. (Figure 3 from Cook and Buja (1997).)}
\label{manipspace}
\end{figure*}
-->

@@

The new approach emerged from experiments on the tour using the linear algebra capabilities, and relatively new interactive graphics interface, available in Mathematica. The components corresponding to $V_m$ are directly controlled by cursor movement, which updates row $m$ of $A$. The updated matrix is then orthonormalised.

@@

<!--
Note: I removed the angular bracets since cdot denotes scalar product already (also used like this below for the slice definitions)
-->

This algorithm will produce the changes to a projection as illustrated in Figure \ref{fig:manualsequence}. The controlled variable, $V_m$, corresponds to the dark line, and sequential changes to row $m$ of $A$ can be seen to roughly follow a specified position (dot). Changes in the other components happen as a result of the orthonormalisation, but are uncontrolled.

@@

```{r manualsequence, out.width="100%", width=12, height=4, fig.cap="Sequence of projections where contribution of one variable is controlled (dark) is changed using unconstrained orthonormalisation. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close."}
source("../code/plot_basis.R")
source("../code/linear_alg.R")

@@

The problem with the new simple method is that it is not faithful to the precise values for $V_m$ because the orthonormalisation will change them. Even though these changes are for the most part imperceptible, one may wish to avoid them and there are numerous ways that this can be enforced, a few are detailed in the Appendix. These primarily differ in how the remaining variables are adjusted during orthonormalisation.

@@

To better explore the space we combine the manual controls for the projection with manual controls for slicing. A slice is a section of the data that is defined by a projection, a center point that is anchoring it in the high-dimensional space and the slice thickness $h$ [@slicetour]. A data point is inside the slice if its orthogonal distance from the projection plane (passing through the center point) is below the thickness $h$. This orthogonal distance is computed in terms of the component that is normal on the projection plane. For $\mathbf{x}_i$ a $p$ dimensional data point and $\mathbf{c}$ the center point (in the same $p$ dimensional space) we compute the orthogonal distance as
\begin{equation}
v_i^2 = ||\mathbf{x}_i' - \mathbf{c}'||^2 = \mathbf{x}_i'^2 + \mathbf{c}'^2 - 2 \mathbf{x}_i'\cdot\mathbf{c}' ,
\label{eq:slice}
\end{equation}
with $\mathbf{c}' = \mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{c}\cdot \mathbf{a}_2 )\mathbf{a}_2$, $\mathbf{x}_i' = \mathbf{x}_i - (\mathbf{x}_i\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{x}_i\cdot \mathbf{a}_2) \mathbf{a}_2$ and $\mathbf{a}_k, k=1,2 (=d)$ denoting the columns of the projection matrix, $\mathbf{A}=(\mathbf{a}_1, \mathbf{a}_2)$.

@@

In addition it is also useful to interactively change the slice thickness $h$ (also called the slice radius), in particular to find the preferred value for exploring the input data. For guidance the estimates of the number of points inside the slice as a function of the original sample size $N$ and the number of dimensions $p$ from @sectionpursuit can be used: in case of a uniform distribution inside a sphere of radius $R$ a slice with thickness $h$ will contain $N_S$ points, with 
\begin{equation}
N_S(h, p, R, N) = \frac{N}{2} \left(\frac{h}{R}\right)^{p-2} \left(p - (p-2)\left(\frac{h}{R}\right)^{2}\right).
\label{eq:count}
\end{equation}

@@

The implementation of the manual tour as suggested here requires the visualization of the current projection in terms of an axis display (see Figure \ref{fig:manualsequence} for an example). This display should be able to track the mouse position and adjust the projection based on the \textcolor{blue}{user's cursor position}. The user can select one axis (corresponding to one variable) by clicking on it, and then adjust the position of that axis by dragging it. In practice, the closest axis will be selected, and the dragging results in tracking of the mouse position which will update the projection in small steps, such that no further interpolation is required.

@@

In addition, we also want to be able to look at slices of the data and select slicing parameters interactively. Here we have implemented a switch to change between projection and slice view, a slider that adjusts the slice thickness, $h$ (named `height` in the Mathematica interface), and a numeric input to specify the center point $c$ explicitly. The display will update based on these input values, in particular the view will jump to a new slice when $c$ is changed.<!-- (though an interpolation might also be useful).-->

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use} `shiny` [@shinym] with `plotly` [@plotly] drawing. However, the `plotly` package is designed to add on to `ggplot2` graphics, not to be a programmable interactive graphics system. It lacks capability to implement the primary tool needed for a manual tour, that is the tracking of the cursor position in small increments. In theory, the historical `locator()` function of base R should provide the necessary capability, but it's behavior is not consistent between graphical devices. The cursor position control is possible with `rgl` [@rgl] but this package is focused on 3D, not the arbitrary dimension of tours. The level of interactivity would have been possible with the `cranvas` system [@cranvas] or `rJava` [@rJava] but neither has a graphics toolkit that currently operates across platforms.  What we have chosen to do is to use the interactivity of Mathematica to experiment, and to implement as an animation in the `tourr` [@tourr] package. There is hope that better interactivity will be available for R soon, perhaps in javascript [e.g. @langevitour, @detourr], which might allow for the  experimentation to develop new techniques.

@@

\textcolor{blue}{The Mathematica implementation relies on the inbuilt function  `LocatorPane`. This function creates a region on the screen where the position of the cursor is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. User interface objects, such as sliders, locator panes, and input fields, are used to provide user control of slice thickness, center, scale and point size.}

@@

Changing slice center manually can be accomplished with the new function, `manual_slice()`. This changes the value of the center point ($\mathbf{c}$) of the slice, along a selected variable axis. It slices from the center of the data out to one side, back to the center, out to the opposite side, and then back to the center. <!--The appendix contains a discussion of a possible visual guide for the slice center location.-->

@@

Potential adjustments for future work include a better visualization of the current center point, potentially with interactive graphics to move the slice, similarly to how the manual tour was implemented here. An alternative would be a step-wise shift of the slice towards the new center point to get a better understanding of the distribution along the selected direction.
-->

@@

```{r penguins-scatmat, out.width="60%", width=10, height=10, fig.align='center', fig.cap="Scatterplot matrix of the (standardised) penguins data. The three species are reasonably different in size, with Gentoo distinguised from the other two on body depth relative to flipper length and body mass."}
library(tidyverse)
library(palmerpenguins)
library(GGally)
penguins_data <- penguins[, c(1,3, 4, 5, 6)] %>%
  drop_na() %>%
  mutate(across(where(is.numeric), function(x) (x-mean(x))/sd(x))) %>%
  rename("bl"="bill_length_mm",
         "bd"="bill_depth_mm",
         "fl"="flipper_length_mm",
         "bm"="body_mass_g")
#ggscatmat(penguins_data, columns = 2:5, col="species") +
# library(paletteer) 
# rcartocolor::ag_Sunset
# rcartocolor::ag_GrnYl
# RColorBrewer::Spectral
# clrs <- c("#245668FF", "#39AB7EFF", "#A9DC67FF")
# clrs <- c("#4B2991FF", "#EA4F88FF", "#F6A97AFF")
# clrs <- c("#4B2991FF", "#39AB7EFF", "#F6A97AFF")
clrs <- c("#245668FF", "#EA4F88FF", "#ABDDA4FF")
penguins_diag <- tibble(x=1, y=c(3,2,1),
     species=levels(penguins_data$species))
pscat <- ggpairs(penguins_data, columns = 2:5, 
          mapping = ggplot2::aes(color = species, 
                                 fill = species),
          upper = list(continuous = "points"),
          diag = list(continuous = "densityDiag")) +
#  scale_colour_manual("", values=c("Adelie" = "#5E82B6", "Chinstrap" = "#E19C24", "Gentoo" = "#8FB132")) +
#  scale_fill_manual("", values=c("Adelie" = "#5E82B6", "Chinstrap" = "#E19C24", "Gentoo" = "#8FB132")) +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  scale_fill_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) 
pscat[1,1] <- ggplot(penguins_diag, 
                     aes(x=x, 
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat[2,2] <- ggplot(penguins_diag, 
                     aes(x=x,
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat[3,3] <- ggplot(penguins_diag, 
                     aes(x=x,
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat[4,4] <- ggplot(penguins_diag, 
                     aes(x=x,
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat + theme_bw() +
  theme(aspect.ratio=1, 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

To illustrate the usefulness of the manual controls we use the 4D penguins data [@penguins] and look at classification models following @sam11271. We will show how classification boundaries can be explored and better understood on projections and slices through 4D space. Figure \ref{fig:penguins-scatmat} shows a scatterplot matrix of this data. There are four variables (`bl = bill_length_mm, bd = bill_depth_mm, fl = flipper_length_mm, bm = body_mass_g`) measuring the size of the penguins from three species (Adelie, Chinstrap and Gentoo). The scatterplot matrix shows that the three species appear to be likely separable, and that at least the Gentoo can be distinguished from the other two species when `bd` is paired with `fl` or `bm`. The steps for exploring boundaries in this example are as follows:

@@

We start by exploring the projections of the model prediction. Figure \ref{proj1} summarizes the process. Rotating manually the view, we can visualize the location of predictions for each of the three species, and also get a sense of the difference between the two models. 
To illustrate this difference, we have manually rotated the projection for the RF model (left plot) to identify a view that shows the non-linear but block-type structure that is typical of this type of model. This particular projection ($A_1$) is exported so that it can later be used to show the LDA model (middle plot) and the actual data (right plot). What can be seen is the linearity of the LDA model, where the boundaries are linear and oblique to the variable axes. And, interestingly, this particular projection of the original data shows very distinct clusters of the three species. That means, the obscuring of the boundaries between groups for both of the models is driven by what is happening in the orthogonal space to the plane of the selected projection.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/proj1.pdf}}
%\centerline{\includegraphics[width=0.32\textwidth]{figures/proj1_rf.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_lda.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_data.png}}
\caption{A projection identified using the manual tour, because it shows differences in classification boundaries induced by different models. A RF model produces blocky, rectangular, boundaries, while the LDA model produces oblique linear boundaries. For comparison, in this projection the data shows the three groups are almost perfectly separable.}
\label{proj1}
\end{figure*}

This example can be reproduced with the code in penguins_exploring_manually.nb and a run through is shown in the video at https://vimeo.com/747585410.

@@

Where else might it be useful to use mathematica

>>>>

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the contribution of a given variable to a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the center of a slice, for example, to explore how structure changes in local subspaces. The Mathematica package as well as example notebooks are provided, which contain functions enabling the user to experiment with these new manual controls, with one specifically for exploring regions and boundaries produced by classification models. The advantage of Mathematica is its linear algebra capabilities and interactive cursor location controls. Some limited implementation has also been made available in the R package \tt{tourr}. 
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}\doublespacing
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, \textcolor{blue}{methods which work for arbitrary dimensions have more general utility}. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional data with low-dimensional projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. (Various videos of tours, e.g. @dataviewer, @gt-pp-video, on the @ASA22, illustrate this concept.) Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video [@PRIM9-video] illustrating the capabilities is available through \textcolor{blue}{the} video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allow one to quickly see many, many different projections. After Asimov, there have been many tour developments, which are summarized in @lee2021.

<!-- Manual tours -->
One such direction of work develops the ideas from PRIM-9, to provide manual control of a tour. @cook_manual_1997 describe controls for 1D (or 2D) projections, respectively in a 2D (or 3D) manipulation space, allowing the user to select any variable axis, and rotate it into, out of, or around the projection through horizontal, vertical, oblique, radial or angular changes in value. @spyrison_spinifex_2020 refined this algorithm and implemented it to generate radial tour animation sequences.

<!-- Purpose -->
Manual controls are especially useful for assessing the sensitivity of structure to particular elements of the projection. There are many places where it is useful. In exploratory data analysis, where one sees clusters in a projection, one may ask whether some variables can be removed from the projection without affecting the clustering. For interpreting models, one can reduce or increase a variable's contribution, to examine the variable importance. Having the user interact with a projection is extremely valuable for understanding high-dimensional data. However, these algorithms have two problems: (1) the pre-processing of creating a manipulation space overly complicates the algorithm, and (2) extending to higher dimensional control is difficult.

Another potentially useful manual control is to allow the user to choose the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the position of the center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center. It would also allow the user to interactively decide on the thickness of the slice.

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods is illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a Mathematica [@Mathematica] package that is used for the application and describes the interactive environment that would be desirable within R [@rref] as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and three Mathematica notebooks that can be used to reproduce the applications.

@@

The methods for updating component values in @cook_manual_1997 (and utilized in @spyrison_spinifex_2020) are prescribed primarily for a 2D projection, to take advantage of (then) newly developed 3D trackball controls made available for computer gaming. The first step was to construct a 3D manipulation space from a 2D projection. In this space, the coefficient of the controlled variable ranges between -1 and 1. Movements of a cursor are recorded and converted into changes in the values of $V_m$ thus changing the displayed 2D projection. The algorithm also provided constraints to horizontal, vertical, radial, or angular motions only. The construction of the manipulation space overly complicates the manual controls, especially when considering possible techniques that will apply to arbitrary $d$.

<!--
\begin{figure*}[ht]
\centerline{\includegraphics[width=0.8\textwidth]{figures/manip_space.pdf}}
\caption{Original construction of the manual tour designed for 2D projections and created a 3D space from which to utilize track ball controls to change its contribution. (Figure 3 from Cook and Buja (1997).)}
\label{manipspace}
\end{figure*}
-->

@@

The new approach emerged from experiments on the tour using linear algebra capabilities, and a relatively new interactive graphics interface, available in Mathematica. The components corresponding to $V_m$ are directly controlled by cursor movement, which updates row $m$ of $A$. The updated matrix is then orthonormalized.

@@

<!--
Note: I removed the angular brackets since cdot denotes scalar product already (also used like this below for the slice definitions)
-->

This algorithm will produce the changes to a projection as illustrated in Figure \ref{fig:manualsequence}. The controlled variable, $V_m$, corresponds to the dark line, and sequential changes to row $m$ of $A$ can be seen to roughly follow a specified position (dot). Changes in the other components happen as a result of the orthonormalization but are uncontrolled.

@@

```{r manualsequence, out.width="100%", width=12, height=4, fig.cap="Sequence of projections where the contribution of one variable is controlled (dark) is changed using unconstrained orthonormalization. The dot indicates the chosen values for the controlled variable, $V_m$. It can be seen that the actual axis does not precisely match the chosen position, but it is close."}
source("../code/plot_basis.R")
source("../code/linear_alg.R")

@@

The problem with the new simple method is that it is not faithful to the precise values for $V_m$ because the orthonormalization will change them. Even though these changes are for the most part imperceptible, one may wish to avoid them and there are numerous ways that this can be enforced, a few are detailed in the Appendix. These primarily differ in how the remaining variables are adjusted during orthonormalization.

@@

To better explore the space we combine the manual controls for the projection with manual controls for slicing. A slice is a section of the data that is defined by a projection, a center point that is anchoring it in the high-dimensional space, and the slice thickness $h$ [@slicetour]. A data point is inside the slice if its orthogonal distance from the projection plane (passing through the center point) is below the thickness $h$. This orthogonal distance is computed in terms of the component that is normal on the projection plane. For $\mathbf{x}_i$ a $p$ dimensional data point and $\mathbf{c}$ the center point (in the same $p$ dimensional space) we compute the orthogonal distance as
\begin{equation}
v_i^2 = ||\mathbf{x}_i' - \mathbf{c}'||^2 = \mathbf{x}_i'^2 + \mathbf{c}'^2 - 2 \mathbf{x}_i'\cdot\mathbf{c}' ,
\label{eq:slice}
\end{equation}
with $\mathbf{c}' = \mathbf{c} - (\mathbf{c}\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{c}\cdot \mathbf{a}_2 )\mathbf{a}_2$, $\mathbf{x}_i' = \mathbf{x}_i - (\mathbf{x}_i\cdot \mathbf{a}_1) \mathbf{a}_1 - (\mathbf{x}_i\cdot \mathbf{a}_2) \mathbf{a}_2$ and $\mathbf{a}_k, k=1,2 (=d)$ denoting the columns of the projection matrix, $\mathbf{A}=(\mathbf{a}_1, \mathbf{a}_2)$.

@@

In addition, it is also useful to interactively change the slice thickness $h$ (also called the slice radius), in particular, to find the preferred value for exploring the input data. For guidance the estimates of the number of points inside the slice as a function of the original sample size $N$ and the number of dimensions $p$ from @sectionpursuit can be used: in case of a uniform distribution inside a sphere of radius $R$ a slice with thickness $h$ will contain $N_S$ points, with 
\begin{equation}
N_S(h, p, R, N) = \frac{N}{2} \left(\frac{h}{R}\right)^{p-2} \left(p - (p-2)\left(\frac{h}{R}\right)^{2}\right).
\label{eq:count}
\end{equation}

@@

The implementation of the manual tour as suggested here requires the visualization of the current projection in terms of an axis display (see Figure \ref{fig:manualsequence} for an example). This display should be able to track the mouse position and adjust the projection based on the \textcolor{blue}{user's cursor position}. The user can select one axis (corresponding to one variable) by clicking on it, and then adjust the position of that axis by dragging it. In practice, the closest axis will be selected, and the dragging results in tracking the mouse position which will update the projection in small steps, such that no further interpolation is required.

@@

In addition, we also want to be able to look at slices of the data and select slicing parameters interactively. Here we have implemented a switch to change between projection and slice view, a slider that adjusts the slice thickness, $h$ (named `height` in the Mathematica interface), and a numeric input to specify the center point $c$ explicitly. The display will update based on these input values, in particular, the view will jump to a new slice when $c$ is changed.<!-- (though an interpolation might also be useful).-->

\textcolor{blue}{Within R, the natural implementation for an interactive interface would use} `shiny` [@shinym] with `plotly` [@plotly] drawing. However, the `plotly` package is designed to add to `ggplot2` graphics, not to be a programmable interactive graphics system. It lacks the capability to implement the primary tool needed for a manual tour, which is the tracking of the cursor position in small increments. In theory, the historical `locator()` function of base R should provide the necessary capability, but its behavior is not consistent between graphical devices. The cursor position control is possible with `rgl` [@rgl] but this package is focused on 3D, not the arbitrary dimension of tours. The level of interactivity would have been possible with the `cranvas` system [@cranvas] or `rJava` [@rJava] but neither has a graphics toolkit that currently operates across platforms.  What we have chosen to do is to use the interactivity of Mathematica to experiment, and to implement the resulting ideas as an animation in the `tourr` [@tourr] package. There is hope that better interactivity will be available for R soon, perhaps in javascript [e.g. @langevitour, @detourr], which might allow for the experimentation to develop new techniques.

@@

\textcolor{blue}{The Mathematica implementation relies on the inbuilt function  `LocatorPane`. This function creates a region on the screen where the position of the cursor is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. User interface objects, such as sliders, locator panes, and input fields, are used to provide user control of slice thickness, center, scale, and point size.}

@@

Changing the slice center manually can be accomplished with the new function, `manual_slice()`. This changes the value of the center point ($\mathbf{c}$) of the slice, along a selected variable axis. It slices from the center of the data out to one side, back to the center, out to the opposite side, and then back to the center. <!--The appendix contains a discussion of a possible visual guide for the slice center location.-->

@@

Potential adjustments for future work include better visualization of the current center point, potentially with interactive graphics to move the slice, similar to how the manual tour was implemented here. An alternative would be a step-wise shift of the slice towards the new center point to get a better understanding of the distribution along the selected direction.
-->

@@

```{r penguins-scatmat, out.width="60%", width=10, height=10, fig.align='center', fig.cap="Scatterplot matrix of the (standardized) penguins data. The three species are reasonably different in size, with Gentoo distinguished from the other two on body depth relative to flipper length and body mass."}
library(tidyverse)
library(palmerpenguins)
library(GGally)
penguins_data <- penguins[, c(1,3, 4, 5, 6)] %>%
  drop_na() %>%
  mutate(across(where(is.numeric), function(x) (x-mean(x))/sd(x))) %>%
  rename("bl"="bill_length_mm",
         "bd"="bill_depth_mm",
         "fl"="flipper_length_mm",
         "bm"="body_mass_g")
#ggscatmat(penguins_data, columns = 2:5, col="species") +
# library(paletteer) 
# rcartocolor::ag_Sunset
# rcartocolor::ag_GrnYl
# RColorBrewer::Spectral
# clrs <- c("#245668FF", "#39AB7EFF", "#A9DC67FF")
# clrs <- c("#4B2991FF", "#EA4F88FF", "#F6A97AFF")
# clrs <- c("#4B2991FF", "#39AB7EFF", "#F6A97AFF")
clrs <- c("#245668FF", "#EA4F88FF", "#ABDDA4FF")
penguins_diag <- tibble(x=1, y=c(3,2,1),
     species=levels(penguins_data$species))
pscat <- ggpairs(penguins_data, columns = 2:5, 
          mapping = ggplot2::aes(color = species, 
                                 fill = species),
          upper = list(continuous = "points"),
          diag = list(continuous = "densityDiag")) +
#  scale_colour_manual("", values=c("Adelie" = "#5E82B6", "Chinstrap" = "#E19C24", "Gentoo" = "#8FB132")) +
#  scale_fill_manual("", values=c("Adelie" = "#5E82B6", "Chinstrap" = "#E19C24", "Gentoo" = "#8FB132")) +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  scale_fill_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) 
pscat[1,1] <- ggplot(penguins_diag, 
                     aes(x=x, 
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat[2,2] <- ggplot(penguins_diag, 
                     aes(x=x,
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat[3,3] <- ggplot(penguins_diag, 
                     aes(x=x,
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat[4,4] <- ggplot(penguins_diag, 
                     aes(x=x,
                         y=y, 
                         label=species, 
                         colour=species)) + 
  geom_text() +
  scale_colour_manual("", values=c("Adelie" = clrs[1], "Chinstrap" = clrs[2], "Gentoo" = clrs[3])) +
  ylim(c(0,4))
pscat + theme_bw() +
  theme(aspect.ratio=1, 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

To illustrate the usefulness of the manual controls we use the 4D penguins data [@penguins] and look at classification models following @sam11271. We will show how classification boundaries can be explored and better understood on projections and slices through 4D space. Figure \ref{fig:penguins-scatmat} shows a scatterplot matrix of this data. There are four variables (`bl = bill_length_mm, bd = bill_depth_mm, fl = flipper_length_mm, bm = body_mass_g`) measuring the size of the penguins from three species (Adelie, Chinstrap, and Gentoo). The scatterplot matrix shows that the three species appear to be likely separable and that at least the Gentoo can be distinguished from the other two species when `bd` is paired with `fl` or `bm`. The steps for exploring boundaries in this example are as follows:

@@

We start by exploring the projections of the model prediction. Figure \ref{proj1} summarizes the process. Rotating manually the view, we can visualize the location of predictions for each of the three species, and also get a sense of the difference between the two models. 
To illustrate this difference, we have manually rotated the projection for the RF model (left plot) to identify a view that shows the non-linear but block-type structure that is typical of this type of model. This particular projection ($A_1$) is exported so that it can later be used to show the LDA model (middle plot) and the actual data (right plot). What can be seen is the linearity of the LDA model, where the boundaries are linear and oblique to the variable axes. And, interestingly, this particular projection of the original data shows very distinct clusters of the three species. That means the obscuring of the boundaries between groups for both of the models is driven by what is happening in the orthogonal space to the plane of the selected projection.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.95\textwidth]{figures/proj1.pdf}}
%\centerline{\includegraphics[width=0.32\textwidth]{figures/proj1_rf.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_lda.png}
%\includegraphics[width=0.32\textwidth]{figures/proj1_data.png}}
\caption{A projection identified using the manual tour because it shows differences in classification boundaries induced by different models. An RF model produces blocky, rectangular, boundaries, while the LDA model produces oblique linear boundaries. For comparison, in this projection, the data shows the three groups are almost perfectly separable.}
\label{proj1}
\end{figure*}

This example can be reproduced with the code in penguins_exploring_manually.nb and a run-through is shown in the video at https://vimeo.com/747585410.

@@

Where else might it be useful to use Mathematica
